---
title: "The LEMIS Database"
author: "Noam Ross, Allison White, Carlos Zambrana-Torrelio"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    keep_md: true
  github_document:
    toc: true
    html_preview: false
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The **lemis** package provides access U.S. Fish and Wildlife Service
data on wildlife and wildlife product imports to and exports from the
United States.  This data was obtained via more than 14 years of FOIA
requests by EcoHealth Alliance.

# Installation

Install the **lemis** package with this command:

```{r install_me, eval = FALSE}
source("https://install-github.me/ecohealthalliance/lemis")
```

As LEMIS currently lives in a private repository, you must have a
GitHub personal access token set up to install and use the package. Instructions
for this can be found [here](http://happygitwithr.com/github-pat.html#step-by-step).

# Usage

The main function in **lemis** is `lemis_data()`. This returns the main cleaned
LEMIS database as a **dplyr** tibble.

**lemis** makes use of [**datastorr**](https://github.com/ropenscilabs/datastorr) to manage data download. 
The first time you run `lemis_data()` the package will download the most recent version of the database (~160MB).  Subsequent calls will load the database from storage on your computer.

The LEMIS database is stored as an efficiently compressed [`.fst` file](https://github.com/fstpackage/fst), and loading it loads it a
a [remote dplyr source](https://github.com/krlmlr/fstplyr).  This means that it does not load fully into memory until _after_ `filter()` commands on run, reducing memoery usage. If you wish to manipulate it as a data frame, simply call `dplyr::collect()` to load
it fully into memory.

Note that the full database will be approximately 1 GB in memory.

`lemis_codes()` returns a data frame with descriptions of the codes used by
USFWS in the various columns of `lemis_data()`. This is useful for lookup
or joining with the main data for more descriptive outputs.  The `?lemis_code`
help file also has a searchable table of these codes.

# Data structure

The LEMIS database is a single table listing import and export shipments of wildlife
products.  Here are the field descriptions, which can be found in `lemis_metadata()`.

```{r table1, echo=FALSE, message=FALSE, fig.cap="Table 1. LEMIS database field names"}
knitr::kable(lemis::lemis_metadata())
```

# Tutorial

This script walks through the basic LEMIS functions

```{r tutorial, eval = FALSE}
library(lemis)
library(dplyr)

# Load LEMIS data first time (downloads ~160MB file), returns remote
# dplyr source to database
l <- lemis_data()

rm(l)
# Load again (will NOT download data, it's already on your machine)
l <- lemis_data()

# Load data into memory (~Will take up 1GB RAM)
ll <- collect(l)
ll
rm(ll)

# Filter on-disk and load into memory (smaller!)

l_filtered <- l %>%
  filter(import_export == "I", port = "SS") %>%
  select(-taxa)
l_filtered
ll_filtered %>% collect()
ll_filtered
rm(ll_filtered)

# Get data dictionary and codes
lemis_metadata()
lemis_codes()

# View interactive searchable data dictionary
?lemis_metadata
?lemis_codes

# List versions of LEMIS data on your machine
lemis_versions()
lemis_version_current()

# Remove a downloaded version
lemis_del(lemis_version_current())

# Now list local versions again
lemis_versions()
# List versions of LEMIS data available remotely
lemis_versions(local=FALSE)
```

# Use Cases

Here we show some examples of questions that can be answered with this data.

```{r setup2, message=FALSE, warning=FALSE, results='hide'}
library(lemis)
library(tidyverse)
library(stringi)
theme_set(theme_minimal())
```

***What kind of animals are being imported through Sault Sainte Marie?***

```{r}
sss_port <- # Lookup the port code
  lemis_codes() %>% 
  filter(field == "port", value=="Sault Sainte Marie") %>% 
  pull(code)
lemis_data() %>% 
  filter(port == sss_port, import_export == "I") %>% #Filter before loading the file to disk
  collect() %>% 
  group_by(generic_name) %>% 
  summarise(shipments = n()) %>% 
  arrange(desc(shipments)) %>% 
  head(10) %>% 
ggplot(aes(x = fct_reorder(generic_name, shipments), y = shipments)) +
  geom_col() +
  coord_flip() +
  xlab("Generic Animal Type") + ylab("Shipments via Sault Sainte Marie 2000-2013")
```

Lots o' bears, it seems.

***In what form are _Pteropus_ imported***

```{r}
lemis_data() %>% 
  filter(genus == "pteropus", import_export == "I") %>% 
  collect() %>% 
  group_by(description) %>% 
  summarize(shipments = n()) %>% 
  left_join(filter(lemis_codes(), field=="description"), by = c("description"="code")) %>% 
  arrange(desc(shipments)) %>% 
ggplot(aes(x = fct_reorder(value, shipments), y = shipments)) +
  geom_col() +
  coord_flip() +
  xlab("Description of Pteropus Imports") + ylab("No. Shipments 2000-2013")
  
```

Mostly as scientific specimens.  There are a few clear errors in this.

***Where are live bats imported from?***

As soon as we start dealing with species it's useful to bring in taxonomic
resources.  The [**taxize**](https://github.com/ropensci/taxize) package is a 
great toolbelt for looking up taxonomic data.

```{r}
library(taxize)
bat_genera <- downstream(get_tsn("Chiroptera", rows=1), downto = "Genus", db = "itis") %>% 
  `[[`(1)%>% 
  filter(rankname == "genus") %>% 
  pull(taxonname) %>% 
  stri_trans_tolower()
lemis_data() %>% 
  filter(import_export == "I", description == "LIV", genus %in% bat_genera) %>% 
  group_by(country_origin) %>% 
  summarize(number = sum(quantity)) %>% 
  left_join(filter(lemis_codes(), field=="country"),
            by=c("country_origin"="code")) %>% 
  ggplot(aes(x=fct_reorder(value, number), y=number)) +
  geom_col() +
  coord_flip() +
  xlab("Origin Country") + ylab("No. Live Bats Imported to U.S., 2000-2013")
  
```

**Who is importing these live bats?***

```{r}
lemis_data() %>% 
  filter(import_export == "I", description == "LIV", genus %in% bat_genera) %>% 
  group_by(us_co) %>% 
  summarize(number = sum(quantity)) %>% 
  ggplot(aes(x=fct_reorder(us_co, number), y=number)) +
  geom_col() +
  coord_flip() +
  xlab("U.S. Importer") + ylab("No. Live Bats Imported to U.S., 2000-2013")
```
Hmm, probably some data errors given that there are several coral and aquarium
companies on this list.

***What countries have the highest fraction of seized shipments?***

```{r}
lemis_data() %>% 
  filter(import_export == "I") %>% 
  group_by(country_origin) %>% 
  summarize(frac_seized = sum(disposition == "S")/n()) %>% 
  left_join(filter(lemis_codes(), field=="country"), by=c("country_origin"="code")) %>%
  arrange(desc(frac_seized)) %>% 
  filter(!is.na(value)) %>% 
  head(30) %>% 
  ggplot(aes(x=fct_reorder(value, frac_seized), y=frac_seized)) +
  geom_col() +
  coord_flip() +
  labs(title="Top 30 Origin Countries by Fraction of Wildlife Shipments Seized",
       x="Origin Country", y="Fraction of Shipments Seized, 2000-2013")
```

